The Efficient Handling of Guards in the Design of RPython's Tracing JIT

### 2012-08-13 talk/vmil2012/paper.tex
Changes based on Stephan's remarks

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}
In this paper we describe and analyze how deoptimization works in the context
of tracing just-in-time compilers. What instructions are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

The goal of this paper is to understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points. They need to store
enough information to rebuild the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of RPython's
tracing JIT and explain the rationale behind the design decisions based on the
numbers provided here.

\cfbolz{this paragraph now suddenly \emph{introduces} guards, despite having talked about them already}
The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase special operations,
referred to as \texttt{guards}, are inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after the
optimization pass over the traced and later compiled parts of the benchmarks,
making guards one of the most common types of operations. Many of these guards
fail rarely or not all during execution. There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item An analysis of guards in the context of RPython's tracing JIT to
  substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}

@@

RPython is constructed from two components:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to program such as (if needed) a garbage collector and with some hints
provided by the author a just-in-time compiler.

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing JITs are a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs typically are
mixed-mode execution environments that also contain an interpreter. The
interpreter profiles the executing program and selects frequently executed code
paths to be compiled to machine code. Many tracing JIT compilers focus on
selecting hot loops.

After profiling identifies an interesting
path, tracing is started, recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, which is typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

\begin{figure}
    \input{figures/unopt-log.tex}
    \caption{Unoptimized trace}
    \label{fig:unopt-trace}
\end{figure}

@@

Since tracing linearizes control flow by following one concrete execution,
the full control flow of a program is not observed.
The possible points of deviation from the trace are guard operations
that check whether the same assumptions observed during tracing
still hold during execution.
Similarly, in the case of dynamic languages guards can also encode type
assumptions.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to reconstruct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

To do this reconstruction, it is necessary to take the values of the SSA
variables of the trace and build interpreter stack frames.  Tracing
aggressively inlines functions, therefore the reconstructed state of the
interpreter can consist of several interpreter frames.

If a guard fails often enough, a trace is started from it
to create a trace tree.
When that happens another use case of resume data
is to construct the tracer state.
After the bridge has been recorded and compiled it is attached to the guard.
If the guard fails later, the bridge is executed. Therefore the resume data of
that guard is no longer needed.

@@

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is often useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
often only the information in the top frame changes from one guard to the next.
The other symbolic frames can often just be reused.
The reason for this is that during tracing only the variables
of the currently executing frame can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the frame stack can be reused.

@@

Guards interact with optimizations in various ways.
Most importantly optimizations try to remove as many operations
and therefore guards as possible.
This is done with many classical compiler optimizations.
In particular guards can be removed by subexpression elimination.
If the same guard is encountered a second time in the trace,
the second one can be removed.
This also works if a later guard is weaker
and therefore implied by a earlier guard.

@@

The other important point of interaction between resume data and the optimizer
is RPython's allocation removal optimization~\cite{bolz_allocation_2011}.
This optimization discovers allocations in the trace that create objects
that do not survive long.
An example is the instance of \lstinline{Even} in Figure~\ref{fig:unopt-trace}.
Allocation removal makes resume data more complex.
Since allocations are removed from the trace it becomes necessary
to reconstruct the objects that were not allocated so far when a guard fails.
Therefore the resume data needs to store enough information
to make this reconstruction possible.

Adding this additional information is done as follows.
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The structure of the heap objects that have to be allocated on guard failure
is described by the virtual objects stored in the symbolic frames.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

@@

After optimization the resulting trace is handed to the over platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the to be
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics, operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing even more the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
followed by a guard that checks the result of the operation are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

\todo{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
    \begin{lstlisting}[mathescape]
$b_1$ = int_eq($i_2$, 1)
guard_false($b_1$)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Separated and merged compilation of operations and guards}
  \label{fig:trace-compiled}
\end{figure}

Each guard in the IR has attached to it a list of the IR-variables required to
rebuild the execution state in case the trace is left through
the guard. When a guard is compiled, in addition to the
condition check two things are generated/compiled. First a special data
structure called \emph{low-level resume data} is created that encodes the
information provided by the register allocator about where the values
corresponding to each IR-variable required by the guard will be stored when
execution reaches the code emitted for the corresponding guard. This data
structure stores the data in a compressed manner using an encoding that uses
8bits to store 7bits of information. This encoding is efficient to create and
provides a compact representation of the needed information,
to maintain an acceptable memory profile.

Second a piece of code is generated for each guard that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline. In case the
condition checked in the guard fails
execution jumps to the corresponding trampoline. In the trampoline the pointer to the
\emph{low-level resume data} is loaded and jumps to generic bailout handler, also known as compensation code,
that is used to leave the compiled trace in case of a guard failure.

@@

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend needed to rebuild the state needs to be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

As explained in previous sections, when a specific guard has failed often enough
a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
compiled. When compiling bridges the goal is that future failures of the guards
that led to the compilation of the bridge should execute the bridge without
additional overhead. In particular the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

@@

Once the bridge has been compiled the guard that led to compiling the bridge is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

All figures in this section do not take garbage collection of machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

@@

Tracing JIT compilers only compile the subset of the code executed in a program
that is traced in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the \texttt{resume data} when
compared to the size of the generated machine code and illustrates why it is important to compress the \texttt{resume data} information.

@@

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to a naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently through using the techniques described above. On the other hand
comparing the results to the xz compression which only requires between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal but a trade-off between the required space and the time needed
to build a good compressed representation of the compressed resume data for the
large amount of guards present in the traces.

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail more than 200
times.\footnote{
    The threshold of 200 is rather high. It was picked experimentally to give
    good results for long-running programs.
}
As described before, for guards that fail more than 200 times, a trace
is recorded that starts from the guard. Afterwards the guard is patched so that later
failures execute the new trace instead of jumping to the trampoline and returning to the interpreter. Hence the
numbers presented for guards that fail more than 200 times represent the 200
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the optimized traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
fail more than 200 times. Also of all failing guards a few fail extremely often
and most fail rarely. The results emphasizes that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper we have concentrated on guards, an operation typically found in
tracing just-in-time compilers and used to denote points of possible control
flow divergence in recorded traces.
We described how, based on the observation that guards are a frequent operation
in traces and that they do not fail often, guards have been implemented in the
high and low level components of RPython's tracing JIT compiler.

Finally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions. Our
experiments showed that, as previously assumed, guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bailout. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, while it still produces a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode compressed state representation.
The compressed state representation is reduces the memory footprint of rarely
used data.

Based on the observation that most guards do not fail very often or at all it
would be worth exploring if a more aggressive compression scheme for guards
would be worth the memory saving in contrast to the increased decoding
overhead. Based on the same observation we would like to explore the concept of
LuaJIT's sparse snapshots and its applicability to PyPy.

>>>>

\todo{the introduction needs some work}
\cfbolz{the first two two paragraphs talk about deoptimization, then it
switches to guards. I would say we should only talk about guards in the
beginning}
\bivab{Introduce in a sentence what guards are}
In this paper we describe and analyze how guards, as a concept of tracing just-in-time compilers, work. Explicating what concepts are used in the
intermediate and low-level representation of the JIT instructions and how these
are implemented.

The goal of this paper is to understand the design constraints when
implementing guards. Guards have a runtime cost, they take time to execute. On
the other hand, guards are possible deoptimization points. They need to store
enough information to rebuild the interpreter state.
Based on the informal observation that guards are among the most common
operations in the traces produced by RPython's tracing JIT, our
goal is to present concrete numbers for the frequency and the overhead related
to guards, explain how they are implemented in the different levels of RPython's
tracing JIT and clarify the rationale behind the design decisions based on the
numbers provided here.

\cfbolz{this paragraph now suddenly \emph{introduces} guards, despite having talked about them already}
The operations executed by an interpreter are recorded by the tracing JIT in
case they are frequently executed (this process is described in more detail in
Section \ref{sec:Resume Data}). During the recording phase special operations,
referred to as \texttt{guards}, are inserted into the recorded trace at all
points where the control flow could diverge. As can be seen in
Figure~\ref{fig:guard_percent} guards account for about 14\% to 22\% of the
operations before and for about 15\% to 20\% of the operations after the
optimization pass over the traced and later compiled parts of the benchmarks,
making guards one of the most common types of operations. Many of these guards
fail rarely or not all during execution.

There are several aspects to consider
in the design and optimization of guards, the first aspect is that due to the
large number of guards the memory overhead related to storing the information
needed for deoptimization should be kept low. A second aspect is that
successfully checking guards, i.e. not leaving the compiled trace, – which is
the common case – should be a cheap operation to execute favouring the on-trace
execution speed in contrast to the deoptimization case where the state has to
be rebuilt using the stored information. These constraints and trade-offs are
what make the design and optimization of guards an important and non-trivial
aspect of the low-level design of a tracing just-in-time compiler.

%Section~\ref{sec:Evaluation} presents Figures about the absolute number of
%operations for each benchmark, and the overhead produced by the information
%stored at the different levels for the guards
In this paper we want to substantiate the aforementioned observations and
describe based on them the reasoning behind and the implementation of guards in
RPython's tracing just-in-time compiler. the contributions of this paper are:
\begin{itemize}
  \item An analysis and benchmark of guards in the context of RPython's tracing Jit.
  %An analysis of guards in the context of RPython's tracing JIT to
  %substantiate the aforementioned observation, based on a set of benchmarks,
  \item detailed measurements about the frequency and the
  overhead associated with guards, and
  \item a description about how guards are implemented in the high\-
  and low-level parts of the JIT and describe the rationale behind the design.
\end{itemize}

@@

RPython is constructed from two components:
\begin{itemize}
 \item the language itself
 \item the translation toolchain used to transform RPython programs to executable units
\end{itemize}
The RPython language
is a statically typed object oriented high level language. The language provides
several features such as automatic memory management
and just-in-time compilation. When writing an interpreter using RPython the
programmer only has to write the interpreter for the language she is
implementing.  The second RPython component, the translation toolchain, is used
to transform the program to a low level representations suited to be compiled
and run on one of the different supported target platforms/architectures such
as C, .NET and Java. During the transformation process
different low level aspects suited for the target environment are automatically
added to the program such as (if needed) a garbage collector
and a just-in-time compiler based on hints provided by the author.

\subsection{RPython's Tracing JIT Compilers}
\label{sub:tracing}
Tracing is a technique of just-in-time compilers that generate code by
observing the execution of a program. VMs using tracing JITs typically are
mixed-mode execution environments that also contain an interpreter. The
interpreter profiles the executing program and selects frequently executed code
paths to be compiled to machine code. Many tracing JIT compilers focus on
selecting hot loops.

After profiling identifies an interesting
path, tracing is started thus recording all operations that are executed on this
path. This includes inlining functional calls.
As in most compilers, tracing JITs use an intermediate representation to
store the recorded operations, which is typically in SSA
form~\cite{cytron_efficiently_1991}. Since tracing follows actual execution the
code that is recorded
represents only one possible path through the control flow graph. Points of
divergence from the recorded path are marked with special operations called
\emph{guards}, these operations ensure that assumptions valid during the
tracing phase are still valid when the code has been compiled and is executed.
In the case of dynamic languages, guards are also used to encode type checks
that come from optimistic type specialization by recording the types of
variables seen during tracing.
After a trace has been recorded it is optimized and then compiled to platform
specific machine code.

@@

\begin{figure}
    \input{figures/unopt-log.tex}
    \caption{Unoptimized trace. Numbers on the right represent the line numbers in the original program. Lines marked with -1 represent lines added by JIT}
    \label{fig:unopt-trace}
\end{figure}

@@

Since tracing linearizes control flow by following one concrete execution,
the full control flow of a program is not observed.
The possible points of deviation from the trace are guard operations
that check whether the same assumptions observed while tracing
still hold during execution.
Similarly, in the case of dynamic languages guards can also encode type
assumptions.
In later executions of the trace the guards can fail.
If that happens, execution needs to continue in the interpreter.
This means it is necessary to attach enough information to a guard
to reconstruct the interpreter state when that guard fails.
This information is called the \emph{resume data}.

To do this reconstruction it is necessary to take the values of the SSA
variables of the trace and build interpreter stack frames.  Tracing
aggressively inlines functions, therefore the reconstructed state of the
interpreter can consist of several interpreter frames.

If a guard fails often enough, a trace is started from it
forming a trace tree.
When that happens another use case of resume data
is to construct the tracer state.
After the bridge has been recorded and compiled it is attached to the guard.
If the guard fails later the bridge is executed. Therefore the resume data of
that guard is no longer needed.

@@

The core idea of storing resume data as compactly as possible
is to share parts of the data structure between subsequent guards.
This is useful because the density of guards in traces is so high,
that quite often not much changes between them.
Since resume data is a linked list of symbolic frames
in many cases only the information in the top frame changes from one guard to the next.
The other symbolic frames can often just be reused.
The reason for this is that during tracing only the variables
of the currently executing frame can change.
Therefore if two guards are generated from code in the same function
the resume data of the rest of the frame stack can be reused.

@@

Guards interact with optimizations in various ways.
Using many classical compiler optimizations the JIT tries to remove as many
operations, and therefore guards, as possible.
In particular guards can be removed by subexpression elimination.
If the same guard is encountered a second time in the trace,
the second one can be removed.
This also works if a later guard is weaker
and hence implied by a earlier guard.

@@

The other important point of interaction between resume data and the optimizer
is RPython's allocation removal optimization~\cite{bolz_allocation_2011}.
This optimization discovers allocations in the trace that create objects
that do not survive long.
An example is the instance of \lstinline{Even} in Figure~\ref{fig:unopt-trace}.
Allocation removal makes resume data more complex.
Since allocations are removed from the trace it becomes necessary
to reconstruct the objects that were not allocated so far when a guard fails.
Consequently the resume data needs to store enough information
to make this reconstruction possible.

Adding this additional information is done as follows:
So far, every variable in the symbolic frames
contains a constant or an SSA variable.
After allocation removal the variables in the symbolic frames can also contain
``virtual'' objects.
These are objects that were not allocated so far,
because the optimizer removed their allocation.
The structure of the heap objects that have to be allocated on guard failure
is described by the virtual objects stored in the symbolic frames.
To this end, the content of every field of the virtual object is described
in the same way that the local variables of symbolic frames are described.
The fields of the virtual objects can therefore be SSA variables, constants
or other virtual objects.
They are encoded using the same compact two-byte representation
as local variables.

@@

After optimization the resulting trace is handed over to the platform specific
backend to be compiled to machine code. The compilation phase consists of two
passes over the lists of instructions, a backwards pass to calculate live
ranges of IR-level variables and a forward pass to emit the instructions. During
the forward pass IR-level variables are assigned to registers and stack
locations by the register allocator according to the requirements of the
emitted instructions.  Eviction/spilling is performed based on the live range
information collected in the first pass. Each IR instruction is transformed
into one or more machine level instructions that implement the required
semantics. Operations without side effects whose result is not used are not
emitted. Guards instructions are transformed into fast checks at the machine
code level that verify the corresponding condition.  In cases the value being
checked by the guard is not used anywhere else the guard and the operation
producing the value can often be merged, further reducing even more the overhead of the guard.
Figure \ref{fig:trace-compiled} shows how an \texttt{int\_eq} operation
followed by a guard that checks the result of the operation are compiled to
pseudo-assembler if the operation and the guard are compiled separated or if
they are merged.

\todo{Figure needs better formatting}
\begin{figure}[ht]
  \noindent
  \centering
  \begin{minipage}{1\columnwidth}
    \begin{lstlisting}[mathescape]
$b_1$ = int_eq($i_2$, 1)
guard_false($b_1$)
    \end{lstlisting}
  \end{minipage}
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
MOVEQ r8, #1
MOVNE r8, #0
CMP r8, #0
BEQ <bailout>
    \end{lstlisting}
  \end{minipage}
  \hfill
  \begin{minipage}{.40\columnwidth}
    \begin{lstlisting}
CMP r6, #1
BNE <bailout>
...
...
...
    \end{lstlisting}
  \end{minipage}
  \caption{Result of separated (left) and merged (right) compilation of operations and guards (top).}
  \label{fig:trace-compiled}
\end{figure}

Attached to each guard in the IR is a list of the IR-variables required to
rebuild the execution state in case the trace is left through
the guard. When a guard is compiled, in addition to the
condition check two things are generated/compiled.

First a special data
structure called \emph{low-level resume data} is created. This data structure encodes the
information about where, i.e. which register or stack location, the IR-variables required to rebuild the state will be stored when the guard is executed.
This data
structure stores the values in a succinct manner using an encoding that uses
8 bits to store 7 bits of information, ignoring leading zeros. This encoding is efficient to create and
provides a compact representation of the needed information,
to maintain an acceptable memory profile.

Second a piece of code is generated for each guard that acts as a trampoline.
Guards are implemented as a conditional jump to this trampoline in case the
guard checked fails.
In the trampoline the pointer to the
\emph{low-level resume data} is loaded and execution jumps to a generic bailout handler, also known as compensation code,
that is used to leave the compiled trace in case of a guard failure.

@@

As in previous sections the underlying idea for the design of guards is to have
a fast on-trace profile and a potentially slow one in the bailout case where
the execution takes one of the side exits due to a guard failure. At the same
time the data stored in the backend, required to rebuild the state, should be as
compact as possible to reduce the memory overhead produced by the large number
of guards, the numbers in Figure~\ref{fig:backend_data} illustrate that the
compressed encoding currently has about 15\% to 25\% of the size of of the
generated instructions on x86.

As explained in previous sections, when a specific guard has failed often enough
a new trace, referred to as a \emph{bridge}, starting from this guard is recorded and
compiled.
Since the goal of compiling bridges is to improve execution speed on the
diverged path (failing guard) they should not introduce additional overhead.
In particular the failure of the guard should not lead
to leaving the compiled code prior to execution the code of the bridge.

@@

Once the bridge has been compiled the corresponding guard is
patched to redirect control flow to the bridge in case the check fails. In
the future, if the guard fails again it jumps to the code compiled for the bridge
instead of bailing out. Once the guard has been compiled and attached to the
loop the guard becomes just a point where control-flow can split. The loop
after the guard and the bridge are just conditional paths.
Figure~\ref{fig:trampoline} shows a diagram of a compiled loop with two guards,
Guard \#1 jumps to the trampoline, loads the \texttt{low level resume data} and
then calls the bailout handler, whereas Guard \#2 has already been patched
and directly jumps to the corresponding bridge. The bridge also contains two
guards that work based on the same principles.
\begin{figure}
\centering
\includegraphics[width=0.5\textwidth]{figures/loop_bridge.pdf}
\caption{Trace control flow in case of guard failures with and without bridges}
\label{fig:trampoline}
\end{figure}
% section Guards in the Backend (end)

@@

All measurements presented in this section do not take garbage collection of machine code into account. Pieces
of machine code can be globally invalidated or just become cold again. In both
cases the generated machine code and the related data is garbage collected. The
figures show the total amount of operations that are evaluated by the JIT and
the total amount of code and data that is generated from the optimized traces.

@@

Tracing JIT compilers only compile the subset of the code executed in a program
that occurs in a hot loop, for this reason the amount of generated machine
code will be smaller than in other juts-in-time compilation approaches.  This
creates a larger discrepancy between the size of the \texttt{resume data} when
compared to the size of the generated machine code and illustrates why it is important to compress the \texttt{resume data} information.

@@

The results show that the current approach of compression and data sharing only
requires 18.3\% to 31.1\% of the space compared to a naive approach. This
shows that large parts of the resume data are redundant and can be stored more
efficiently through using the techniques described above. On the other hand
comparing the results to the xz compression which only needs between 17.1\%
and 21.1\% of the space required by our compression shows that the compression
is not optimal but a trade-off between the required space and the time needed
to build a good compressed representation of the compressed resume data for the
large amount of guards present in the traces.

\subsection{Guard Failures}
\label{sub:guard_failure}
The last point in this discussion is the frequency of guard failures.
Figure~\ref{fig:failing_guards} presents for each benchmark a list of the
relative amounts of guards that ever fail and of guards that fail often enough that a bridge is compiled.
\footnote{
    The threshold used is 200 failures. This rather high threshold was picked experimentally to give
    good results for long-running programs.
}

After the guard is patched
failures execute the new bridge instead of jumping to the trampoline and returning to the interpreter. Hence the
numbers presented for guards that have a bridge represent the
failures up to the compilation of the bridge and all executions of the then
attached bridge.

@@

From Figure~\ref{fig:failing_guards} we can see that only a very small amount
of all the guards in the optimized traces ever fail. This amount varies between
2.4\% and 5.7\% of all guards. As can be expected, even fewer guards fail often
enough that a bride is compiled for them, only 1.2\% to 3.6\% of all guards
fail often enough that a bridge is compiled. Also of all failing guards a few fail extremely often
and most fail rarely. The results emphasizes that as most of the guards never
fail it is important to make sure that the successful execution of a guard does
not have unnecessary overhead.

@@

\section{Conclusion}
\label{sec:Conclusion}
In this paper we have concentrated on guards, an operation typically found in
tracing just-in-time compilers and used to denote points of possible control
flow divergence in recorded traces.
Based on the observation that guards are a frequent operation in traces and
that they do not fail often, we described how they have been implemented in the
high and low level components of RPython's tracing JIT compiler.

Finally we have presented experimental data collected using the standard PyPy
benchmark set to evaluate previous observations and assumptions. Our
experiments confirmed that guards are a very common
operation in traces. At the same time guards are associated with a high
overhead, because for all compiled guards information needs to be
stored to restore the execution state in case of a bailout. The measurements
showed that the compression techniques used in PyPy effectively reduce the
overhead of guards, but it still produces a significant overhead. The results
also showed that guard failure is a local event: there are few
guards that fail at all, and even fewer that fail very often.
These numbers validate the design decision of reducing the overhead of
successful guard checks as much as possible while paying a higher price in the
case of bailout due to having to decode compressed state representation.
The compressed state representation reduces the memory footprint of rarely
used data.

Based on the observation that guard failure is rare it
would be worth exploring if a more aggressive compression scheme for guards
would be worth the memory saving in contrast to the increased decoding
overhead. Based on the same observation we would like to explore the concept of
LuaJIT's sparse snapshots and its applicability to PyPy.
